Ecological performance of detecting data fabrication

### 2017-07-25 submission/manuscript.Rmd
Update manuscript.Rmd

The application of such statistical methods to detect data fabrication has occurred in several cases of scientific misconduct in recent years and has potential for future application beyond these specific cases. For example, problems in papers by Fuji were highlighted with statistical methods [@10.1111/j.1365-2044.2012.07128.x;@10.1111/anae.13126], resulting in 183 retractions [@oransky2015]. This method works as follows: in true randomized clinical trials (RCTs), baseline measurements should be statistically equal across groups. As such, the $p$-values for group comparisons would be expected to be uniformly distributed because the null hypothesis of equal groups is true by definition of the design. In the Fuji papers, group comparisons showed excessive consistency, resulting primarily in high $p$-values (e.g., .99, .95) and a high mean $p$-value across the comparisons, where a mean $p$-value of .5 is expected. As an illustration, see Table `r tabnr`, which depicts the assignment of participants to the experimental or control condition for a true randomized design (Set 1) or for a fabricated design (Set 2). The mean $p$-value for the true randomized design Set 1 is `r round(mean(ex1p), 3)`, whereas the fabricated Set 2 has mean $p$-value `r round(mean(ex2p), 3)`. Other statistical methods to detect data fabrication are addressed in the theoretical framework.
.

@@

When the $p$-value distribution is not uniform nor right-skewed, it can indicate potential data anomalies. For example, a distribution might show a bump when excessive $p$-hacking occurs that seeks barely significant results [e.g., Figure 1 in @10.7717/peerj.1935]. However, data fabricators might create other non-uniform or non-right-skewed distributions, failing to take into account the theoretical boundary conditions. For example, given the frequent misinterpretation of $p$-values [@10.3758/BF03213921;@10.1007/s11336-015-9446-0;@10.7326/0003-4819-130-12-199906150-00008], it is possible that the same underlying misinterpration will manifest itself in fabricated data. More specifically, given the frequent misinterpretation of $p$-valuesas the probability of an effect [@10.1053/j.seminhematol.2008.04.003;@10.1136/bmj.311.7003.485], researchers who fabricate nonsignificant data might forget to fabricate a uniform $p$-value distribution and generate only large $p$-values when fabricating nonsignificant results.

In order to test whether observed $p$-values are anomalous, we proposed an adaptation of Fisher's method [@10.1186/s41073-016-0012-9]. This adaptation is a simple reversal of the original Fisher method [@fisher1925], which was introduced as a meta-analytic test for the presence of an effect. This test is computed as 
$$
\chi^2_{2k}=-2\sum\limits^k_{i=1}\ln(p_i)
$$
where it tests for right-skew (i.e., more smaller $p$-values than larger $p$-values) across the $k$ number of $p$-values. Reversing the Fisher method results in 
$$
\chi^2_{2k}=-2\sum\limits^k_{i=1}\ln(1-\frac{p_i-t}{1-t})
$$
where it now tests for left-skew (i.e., more larger $p$-values than smaller $p$-values) across the $k$ number of $p$-values that falls above the threshold $t$. This threshold is added and can be set to any value between zero and one. Upon writing this paper, it became clear to us that this is reversed Fisher method is similar to the operating principle of Carlisle's method testing for excessive homogeneity across baseline measurements in RCTs [@10.1111/anae.13938;@10.1111/j.1365-2044.2012.07128.x;@10.1111/anae.13126].

@@

Terminal digit analysis is based on the principle that the rightmost digit is the most random digit of a number, hence, is expected to be uniformly distributed under specific conditions [@10.1080/08989629508573866;@10.1080/03610919608813325]. Terminal digit analysis is conducted with a $\chi^2$-test ($df=10-1$) on the digit occurrence counts (including zero), where the observed frequencies are compared with the expected uniform frequencies. Tthe rule of thumb outlined by @agresti2003categorical indicates at least 50 observations are required to provide a meaningful test of the terminal digit distribution ($n=I\times J \times 5$, with $I$ rows and $J$ columns). Terminal digit analysis was developed during the Imanishi-Kari case by @10.1080/03610919608813325 [for a history of this decade long case, see @kevles2000baltimore].

@@

To test the validity of statistical methods to detect data fabrication in raw data, we investigated raw data of a Stroop experiment [@10.1037/h0054651]. In the Stroop task, participants are asked to determine the color a word is presented in (i.e., word colors), but the word also reads a color (i.e., color words). The presented word color (i.e., 'red', 'blue', or 'green') can be either presented in the congruent color (e.g., 'red' presented in red) or an incongruent color (i.e., 'red' presented in green). The dependent variable in the Stroop task is the response latency (in this study milliseconds are used). Participants in actual studies are typically presented with a set of these, where the mean and standard deviation per condition serves as the raw data. The Stroop effect typically is computed as the difference in mean response latencies between the congruent and incongruent conditions.

>>>>

The application of such statistical methods to detect data fabrication has occurred in several cases of scientific misconduct in recent years and has potential for future application beyond these specific cases. For example, problems in papers by Fuji were highlighted with statistical methods [@10.1111/j.1365-2044.2012.07128.x;@10.1111/anae.13126], resulting in 183 retractions [@oransky2015]. This method works as follows: in true randomized clinical trials (RCTs), baseline measurements should be statistically equal across groups. As such, the $p$-values for group comparisons would be expected to be uniformly distributed because the null hypothesis of equal groups is true by definition of the design. In the Fuji papers, group comparisons showed excessive consistency, resulting primarily in high $p$-values (e.g., .99, .95) and a high mean $p$-value across the comparisons, where a mean $p$-value of .5 is expected. As an illustration, see Table `r tabnr`, which depicts the assignment of participants to the experimental or control condition for a true randomized design (Set 1) or for a fabricated design (Set 2). The mean $p$-value for the true randomized design Set 1 is `r round(mean(ex1p), 3)`, whereas the fabricated Set 2 has mean $p$-value `r round(mean(ex2p), 3)`. Other statistical methods to detect data fabrication are addressed in the theoretical framework.

@@

When the $p$-value distribution is not uniform nor right-skewed, it can indicate potential data anomalies. For example, a distribution might show a bump when excessive $p$-hacking occurs that seeks barely significant results [e.g., Figure 1 in @10.7717/peerj.1935]. However, data fabricators might create other non-uniform or non-right-skewed distributions, failing to take into account the theoretical boundary conditions. For example, given the frequent misinterpretation of $p$-values [@10.3758/BF03213921;@10.1007/s11336-015-9446-0;@10.7326/0003-4819-130-12-199906150-00008], it is possible that the same underlying misinterpration will manifest itself in fabricated data. More specifically, given the frequent misinterpretation of $p$-values as the probability of an effect [@10.1053/j.seminhematol.2008.04.003;@10.1136/bmj.311.7003.485], researchers who fabricate nonsignificant data might forget to fabricate a uniform $p$-value distribution and generate only large $p$-values when fabricating nonsignificant results.

In order to test whether observed $p$-values are anomalous, we proposed an adaptation of Fisher's method [@10.1186/s41073-016-0012-9]. This adaptation is a simple reversal of the original Fisher method [@fisher1925], which was introduced as a meta-analytic test for the presence of an effect. This test is computed as 
$$
\chi^2_{2k}=-2\sum\limits^k_{i=1}\ln(p_i)
$$
where it tests for right-skew (i.e., more smaller $p$-values than larger $p$-values) across the $k$ number of $p$-values. Reversing the Fisher method results in 
$$
\chi^2_{2k}=-2\sum\limits^k_{i=1}\ln(1-\frac{p_i-t}{1-t})
$$
where it now tests for left-skew (i.e., more larger $p$-values than smaller $p$-values) across the $k$ number of $p$-values that falls above the threshold $t$. This threshold is added and can be set to any value between zero and one. Upon writing this paper, it became clear to us that this reversed Fisher method is similar to the operating principle of Carlisle's method testing for excessive homogeneity across baseline measurements in RCTs [@10.1111/anae.13938;@10.1111/j.1365-2044.2012.07128.x;@10.1111/anae.13126].

@@

Terminal digit analysis is based on the principle that the rightmost digit is the most random digit of a number, hence, is expected to be uniformly distributed under specific conditions [@10.1080/08989629508573866;@10.1080/03610919608813325]. Terminal digit analysis is conducted with a $\chi^2$-test ($df=10-1$) on the digit occurrence counts (including zero), where the observed frequencies are compared with the expected uniform frequencies. The rule of thumb outlined by @agresti2003categorical indicates at least 50 observations are required to provide a meaningful test of the terminal digit distribution ($n=I\times J \times 5$, with $I$ rows and $J$ columns). Terminal digit analysis was developed during the Imanishi-Kari case by @10.1080/03610919608813325 [for a history of this decade long case, see @kevles2000baltimore].

@@

To test the validity of statistical methods to detect data fabrication in raw data, we investigated raw data of a Stroop experiment [@10.1037/h0054651]. In the Stroop task, participants are asked to determine the color a word is presented in (i.e., word colors), but the word also reads a color (i.e., color words). The presented word color (i.e., 'red', 'blue', or 'green') can be either presented in the congruent color (e.g., 'red' presented in red) or an incongruent color (i.e., 'red' presented in green). The dependent variable in the Stroop task is the response latency (in this study milliseconds are used). Participants in actual studies are typically presented with a set of these, where the mean and standard deviation per condition serve as the raw data. The Stroop effect typically is computed as the difference in mean response latencies between the congruent and incongruent conditions.
